---
output:
  html_document: default
  pdf_document: default
---
NITRATE CONCENTRATIONS IN SURFACE WATER by Tim Franken
========================================================

# Introduction to the data
The dataset for this project is a datset related to water quality measurements
in small rivers in Flanders. The dataset has 4211 rows of 15 variables:

* nitrate: Yearly maximum nitrate concentration in the river [mg N/L]
* year: Year of measurement
* rainfall: Cumulative rainfall in the month leading up to the measurement [mm]
* oxygen: Oxygen saturation [%]
* area: area of farmland divided by total upstream area [%]. 0 --> no farming,
1 --> all farmland 
* emission: emmission of N by animals divided by area of farmland [kg / m²]
* stockage: stockage of manure in a year [kg / m²]
* grass: area of grass divided by area of farmland
* vegetableG1: area of vegetables in group G1 divided by area of farmland [%]
* vegetableG2: area of vegetables in group G2 divided by area of farmland [%]
* vegetableG3: area of vegetables in group G3 divided by area of farmland [%]
* catchcrop: Factor indicating whether yes or no a catchcrop is planted.
(https://en.wikipedia.org/wiki/Catch_crop)
* norm_N: threshold of N to be applied on the field [kg/m²]
* depostion: depostion of N from the air in the total upstream area [kg/m²]

What we are mainly interested here is to see if any of these variables can
explain the variance in the nitrate concentrations in the river.

First, let's take a look at the general structure of the data and some
summaries.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(GGally)
library(dplyr)
library(memisc)

```

```{r echo=FALSE,  message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
nc <- read.csv('DataSet2.csv', sep=';')
# Change the year variable to a factor
nc$year <- factor(nc$year)
# Some basic information
str(nc)
head(nc)
summary(nc)
```


# Univariate Plots Section

The main variable that we are interested in is nitrate concentrations in the
river. By law there is a threhsold of 11.3 mg N/L.

```{r echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x = nitrate, data=nc, binwidth=1) + 
  geom_segment(aes(x =11, y = 0, xend = 11, yend = 500),  linetype="dashed",
               color="red")

sum(nc$nitrate>11.3) / nrow(nc)

```

As we can see there is a significant number of measurements (25%) which exeed 
the given threshold which is depicted by the dashed red line.

We also want to take a look at the other variables.

```{r echo=FALSE,  message=FALSE, warning=FALSE}
qplot(x=oxygen, data=nc, bins=50)
# Lots of zeros in the database. These might cause problems in subsequent
# analysis. Might be better to delete them.
qplot(x=oxygen, data=subset(nc, oxygen>0), bins=50)
sum(nc$oxygen>0)
sum(nc$oxygen>0) / nrow(nc)
```

Oxygen has a lot of 0 measurements. The plots above show the histogram with and
without the 0 measurements.
These are measurements that are below the detection limit. This is important to
remember once we start building our model.

We still have 2620 measuremnts or about 62% of the total if we would eliminate
those with 0 measurements.

The histograms of the other univariate quantitative variables are given below.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Generate a set of histograms
qplot(x=emission, data=nc)
qplot(x=area, data=nc)
qplot(x=stockage, data=nc, binwidth=0.001)
qplot(x=grass, data=nc)
qplot(x=norm_N, data=nc)


qplot(x=deposition, data=nc)
# Check the difference in both histograms when subsetting for the outliers
qplot(x=deposition, data=subset(nc, deposition < quantile(nc$deposition, 0.99)),
      binwidth=0.001)

# Filter the dataset
nc <- nc %>% filter(deposition < quantile(deposition, 0.99))

```

Based on these histograms we see that depostition contains a couple of outliers.
These are removed from the dataset for further analysis. Apart from this, the
histograms do not indicate clear outliers. It is also clear that not all data
is normally distributed.

Vegetables are divided in 3 different groups containing different types of
vegetables. For this exercise we lump them together in 1 group of vegetables. 
First we check the percentage of data for each group and the total group. That
implies that the value is higher than 0.


```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Get percentage of datapoints with values higher than 0 for all groups
sum(nc$vegetableG1 >0) / nrow(nc)
sum(nc$vegetableG2 >0) / nrow(nc)
sum(nc$vegetableG3 >0) / nrow(nc)

# Create new variable that is the sum of 3 existing variables
nc$total_veg = nc$vegetableG1 + nc$vegetableG2 + nc$vegetableG3

# Get percentage of datapoints with values higher than 0 for all groups
sum(nc$total_veg >0) / nrow(nc)

```

Finally, let's take a look at the resulting histogram, both with and without
the 0 values.

```{r echo=FALSE,  message=FALSE, warning=FALSE}
#Histogram of the new variable, total vegetable.
qplot(x=total_veg, data=nc)

# Histogram without the 0 values.
qplot(x=total_veg, data=subset(nc, total_veg >0))
```


# Univariate Analysis

### What is the structure of your dataset?
The initial dataset consists of 4211 rows and 15 variabales. Each row
consists of a nitrate measurement at a certain location along with some other
variables linked to it.

### What is/are the main feature(s) of interest in your dataset?
The main variable of interest is the nitrate concentrations in the surface
water. This concentration is still too high (related to the allowed threshold)
in a number of locations despite lot of efforts. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
The bivariate analysis might reveal some variables that correlate to the 
overall measured nitrate concentrations. However, since a lot of processes 
determine the final concentrations we don't expect very strong relations.

### Did you create any new variables from existing variables in the dataset?
I created the total_Veg variable which is the sum of the different groups of 
vegetables in the original dataset.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
I removed some outliers in the depostion variable.
Apart from that we need to note that there are a lot of 0 measurements in the
oxygen variable that might influence further analysis.

# Bivariate Plots Section


First we want to investigate the trend in nitrate conccentrations in time. We 
plot different figures which all show similar information.


```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Create historams for each year using facet_wrap
qplot(x = nitrate, data=nc, bins=30) + 
  geom_segment(aes(x =11, y = 0, xend = 11, yend = 100),  linetype="dashed",
               color="red") + 
  facet_wrap(~year, nrow=2)

# Plot the information in one plot using freqpoly.
qplot(x=nitrate, data=nc,
      geom='freqpoly',
      color=year)

# This overview is actually better with just a boxplot.
qplot(x=year, y=nitrate, data=nc, geom = 'boxplot')

# Create new dataframe that contains some summmary statistics per year.
nc.year <- nc %>% group_by(year) %>%
  summarise(median_nitrate = median(nitrate),
            low_quant = quantile(nitrate, 0.05),
            high_quant = quantile(nitrate, 0.95),
            nr_of_exc = sum(nitrate>11.3))

# Create summary of the new dataframe
summary(nc.year)

# plot number of exceedences of the threshold
ggplot(aes(x=year, y=nr_of_exc, group=1), data=nc.year) +
  geom_line()

```

All of the plots show a similar pattern. Since 2010 median nitrate
concentrations have been decreasing. This can be derived from the histograms, 
the frequency polygons as well as the summary values.
Also the number of exceedances (nitrate concentrations > 11.3 mg N/L) has been
decreasing in the last years as can be seen in the last plot.

In order to get a better view on the trend in time we create a summary plot.


```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Create an overview plot of all measurements with some summaries.
# This is similar than the boxplot but connecting lines give a better view.
ggplot(aes(x=year, y=nitrate, group=1), data=nc) + 
  geom_point(alpha=1/30) + 
  geom_point(stat="summary", fun.y = median, color="red") +
  geom_line(stat="summary", fun.y= median, color="red") +
  geom_point(stat="summary", fun.y= quantile, fun.args=list(probs=0.95),
             color="blue") +
  geom_line(stat="summary", fun.y= quantile, fun.args=list(probs=0.95),
            color="blue") + 
  geom_point(stat="summary", fun.y= quantile, fun.args=list(probs=0.05),
             color="blue") +
  geom_line(stat="summary", fun.y= quantile, fun.args=list(probs=0.05),
            color="blue")
```


The red line in this plot shows the median concentration for each year. The
blue lines show the 0.05 & 0.95 quantiles. Although this plot shows similar 
information as the box plot, the decreasing trend is better visualised.

Second question we have is whether or not the use of a catch crop decreases the
observed concentrations in the river

```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Create summary of the dataframe per catch crop
nc.catch <- nc %>% group_by(catchcrop) %>%
  summarise(median_nitrate = median(nitrate),
            low_quant = quantile(nitrate, 0.05),
            high_quant = quantile(nitrate, 0.95))

head(nc.catch)

# Summarize information in freqpoly and boxplot
qplot(x=nitrate, data=nc,
      geom='freqpoly',
      color=catchcrop)

qplot(x=catchcrop, y=nitrate, data=nc, geom = 'boxplot')

```

Both resulting plots and the table do show an effect of the catch crop. Median 
nitrate concentrations are lower if a catch crop is used.

Let's investigate the correlations between some other quantitative variables. 
First we draw a scatter matrix to give a first idea of which variables are
worth investigating further.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Create scatter matrix to get an initial idea of the most important variables
ggpairs(subset(nc, select=c("nitrate", "oxygen", "area", "emission", "stockage",
                            "grass", "norm_N", "deposition",
                            "total_veg")))
```


Based on this information, let's take a closer look at some of the correlations.

Oxygen seems to have a rather high correlation coefficient.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

ggplot(aes(x=oxygen, y=nitrate), data=subset(nc, oxygen>0)) + 
  geom_point(alpha=1/3)

```

The plot does show some correlation although it is not very clear.

As well as area.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

ggplot(aes(x=nitrate, y=area), data=nc) + 
  geom_point(alpha=1/3)

# Break down nitrate concentrations in buckets
nc$nitrate.bucket <- cut(nc$nitrate, breaks=c(0, 5, 10, 15, 20, 25, 30))
# Plot the buckets to get a clearer overview of the realtion between area and N
qplot(x=nitrate.bucket, y=area, data=nc, geom = 'boxplot')

```

From these plots it is clear that there is a correlation between both.
As the first plot was still prone to overplotting I subdived the nitrate in 
different buckets and plotted boxplots for each of them. This gives a clearer
image of the situation.
Higher nitrate concentrations have in general higher areas. As a reminder, area
indicates the proporation of farmland in the upstream area.

Yet it is also clear that a lot of other factors are influencing the 
concentrations.

Area is also strongly linked to deposition which in itself has an inverse
relation to nitrate concentrations. This is rather unexpected as we would 
expect that farm activities have an influence on the depostion.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

ggplot(aes(x=deposition, y=area), data=nc) + 
  geom_point(alpha=1/5)

```

Also the variables emission, grass & total_veg have a small but existing
correlation.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

ggplot(aes(x=emission, y=nitrate), data=nc) + 
  geom_point(alpha=1/5) + 
  xlim(0,0.02)

ggplot(aes(x=grass, y=nitrate), data=nc) + 
  geom_point(alpha=1/5)

ggplot(aes(x=total_veg, y=nitrate), data=subset(nc, total_veg>0)) + 
  geom_point(alpha=1/5)

```

Based on the plot we would assume that the correlation between total_veg &
nitrate is higher that what we observed in the correlation scatter plot. This 
is due to the 0 values in the total_veg variable as can be seen by comparing
the correlation coefficient in both.

```{r echo=FALSE,  message=FALSE, warning=FALSE}

# Investigate and print the corrlation between vegetables and nitrate with and
# without 0 values.
with(nc, cor.test(nitrate, total_veg))
with(subset(nc, total_veg>0), cor.test(nitrate, total_veg))

```

This analysis indicates that it is important to take note of the 0 values in
the subsequent analysis.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Nitrate correlates with quite some of the observed variables. 
yet none of the correlations is particularly strong.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
Yes. There was a high correlation between depostion and area which was rather
unexpected.

### What was the strongest relationship you found?
The strongest relation could be obsvered between nitrate and oxygen. This is
a logical conclusions since oxygen levels will be influence by nitrate 
concentrations.


# Multivariate Plots Section

Before going to our main variable (nitrate) I just want to highlight that
area is also strongly linked to deposition which in itself has an inverse
relation to nitrate concentrations. All of this can be seen in the following
plot.


```{r echo=FALSE,  message=FALSE, warning=FALSE}

ggplot(aes(x=area, y=deposition, color=nitrate), data=nc) + 
  geom_point(alpha=1/5) + 
  scale_color_continuous(low='green', high='red')

```

The correlations between nitrate and other variables are rather week and plots
similar to the above do not really bring much new insights.

Therefore we will try to develop a linear model in this section to predict
nitrate concentrations in the river

```{r echo=FALSE,  message=FALSE, warning=FALSE}

# fit model based on all variables that were identified to have an influence
# in the above.
fit1 <- lm(nitrate ~ oxygen, data = nc)
fit2 <- update(fit1, ~ . + area)
fit3 <- update(fit2, ~ . + emission)
fit4 <- update(fit3, ~ . + grass)
fit5 <- update(fit4, ~ . + total_veg)
fit6 <- update(fit5, ~ . + catchcrop)

# Generate
mtable(fit1, fit2, fit3, fit4, fit5, fit6)

# Let's predict our data based on the model
nc$estimate <- predict(fit5)

# Finally plot the measured values against the predict ones. Add a bissector
# to get a better idea of how far off we are.
ggplot(aes(x=estimate, y=nitrate), data=nc) + 
  geom_point(alpha=1/5) + 
  geom_segment(aes(x = 0, y = 0, xend = 20, yend = 20), linetype="dashed",
               color="red")

```

The last plot shows the measured values against the predicted values based on
the lineair regression model (fit5). We did not include the last variable
(catchcrop) in the equations since it did not result in further improvement of
the model.
This plot indicates that the model is capable of capturing some of the observed
variance but that it is far from perfect.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
We could not find clear variables that strengthened each other based on the
plots. 


### Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

We create a linear model based on the variables oxygen, area, emission, grass &
total_veg This last variable did not bring in additional information and is left
out of the final model.

The R-square of the model is 0.409 which is not great.
The plot with the comparison between measured and predcited concentrations also
indicates that is capable of capturing some of the observed variance but that
it is far from perfect.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
qplot(x = nitrate, data=nc, binwidth=1, color = I('black'), fill = I('#099DD9')) + 
  geom_segment(aes(x =11, y = 0, xend = 11, yend = 500),  linetype="dashed",
               color="red") +
  ggtitle('Histogram of measured nitrate concentrations') + 
  xlab('Nitrate concentration [mg N/L]') +
  ylab('Count[-]')

```

### Description One
This plot forms the basis for this exploration. It shows a histogram of nitrate
concentrations in Flemish rivers, indicating the allowed threshold of
11.3 mg N/L. This plot clearly shows that there is a significant number of
measurement locations that exceed this threshold.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x=nitrate, y=area*100), data=nc) + 
  geom_point(alpha=1/3) + 
  ggtitle('Nitrate concentrations versus proportion of farmland') + 
  xlab('Nitrate concentration [mg N/L]') +
  ylab('Proportion of farmland in upstream area [%]')

```

### Description Two
This plot indicates that there is a relation between the proportion of farmland
in the upstream area and the nitrate concentrations measured in the river. 
This is an important conclusion as it indicates that human behaviour is (partly)
responsible for the measured concentrations. Yet it also becomes clear from
this plot that there are still other influencing factors.


### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x=estimate, y=nitrate), data=nc) + 
  geom_point(alpha=1/5) + 
  geom_segment(aes(x = 0, y = 0, xend = 20, yend = 20), linetype="dashed", 
               size=1.1, color="red") + 
  ggtitle('Measured versus predicted nitrate concentrations') + 
  xlab('Predicted concentration [mg N/L]') +
  ylab('Measured concentration [mg N/L]')

```

### Description Three
This plot shows the comparison between measured and predicted nitrate
concentrations. The predicted concentrations are based upon a linear model that
includes the variables:

* oxygen
* area
* emission
* grass
* total vegetables

This plot shows that we can explain some of the observed variance based on
these variables but definetly not all.


------

# Reflection

This dataset contains 4211 measurments of nitrate and related variables in 
rivers in Flanders. About 25% of the measurement locations exceeds the 
prescribed threshold of 11.3 mg N/L. Based on a (rudimentary) trend analysis we
can conclude that this number is decreasing over the past years.

We can conclude from the bivariate analysis that nitrate concentrations
correlate well with oxygen in the water and the proportion of farmland in the
upstream are. Also other variables such as the area of grassland, vegetable and 
emmission of N by animals is correlated with nitrate concentrations.

Furthermore we found that a large proportion of the data contains a high number
of 0 values in the dataset. This is important to take into account in further
investigations.

The linear model we set up indicated that we can explain some of the observed
variance based on the variables in our dataset but definetly not all, which
could be further investigated in future work.

The dataset we used still contains a limited number of variables and does e.g. 
not contain any information related to physical or chemical properties of the 
soil such as e.g. soil texture which might influence nitrate concentrations as
well. For future work it might be worth including such variables in the
analysis.

